Gusto: a toolkit for compatible finite element dynamical cores

### 2025-04-03 source/paper.tex
add some comments to sec1

Compatible finite elements provide a route to stable, consistent
discretisations of the partial differential equations governing
geophysical fluids which also have the correct wave propagation speeds
and avoid spurious computational modes, even on non-uniform
grids. These properties, along with conservation of mass and
potentially other quantities such as energy and enstrophy, are
considered essential for the accuracy of the dynamical core of weather
and climate models \citep{staniforth2012horizontal}. Achieving these
properties on a non-orthogonal grid is important because of parallel
scaling**. While the required properties of the discretisation guide the
choice of finite element spaces, there are many options for both the
spatial and temporal discretisation and it is not clear which will be
best for any particular problem. Gusto provides a flexible,
extensible, easy-to-use toolkit of options for rapid prototyping of
different spatial and temporal discretisations based on the compatible
finite element framework applied to geophysical fluid dynamics
equations relevant to numerical weather and climate prediction. In
this paper we introduce Gusto to the community, summarise its current
capabilities and outline our vision for future development.

@@

The finite element method represents fields (e.g. velocity, pressure,
density or temperature) on a mesh made of non-overlapping, conforming
cells, or elements. Within each element, the field is represented as a
sum of coefficients multiplied by polynomial basis functions. The
finite element space is specified by the degree of the polynomial
basis functions and the continuity of these functions between the
elements. The partial differential equations used in weather and
climate modelling have multiple prognostic fields and it is not
necessary, or indeed desirable, to use the same finite element
function space for each field. Using different finite element function
spaces for each field gives what is known as a mixed finite element
discretisation.

@@

The compatible finite element spaces relevant for geophysical fluid
dynamics are constructed from a discrete differential complex called
the de Rham complex. This is a sequence of spaces with differential
operators between them that map onto the kernal of the next operator
in the sequence. This is often summarised in a commutation diagram as
in figures \ref{fig:2D de Rham} and \ref{fig:3D de Rham}. In these
diagrams the infinite dimensional spaces are along the top row and the
$\pi_i$ denote projections into the corresponding finite dimensional
subspace used in the compatible finite element discretisation. The
differential operators are denoted by $\diff^k$ and these commute with
the projections, meaning that $\diff^{k+1}\pi_k(\cdot) =
\pi_{k+1}\diff^{k+1}(\cdot)$. This guarantees that, in 3D, $\nabla$
maps onto the kernal of $\nabla\times$, ensuring
$\nabla\times\nabla(\cdot)=0$, and $\nabla\times$ maps onto the kernal
of $\nabla\cdot$, ensuring that $\nabla\cdot\nabla\times(\cdot)=0$;
i.e. the spaces are compatible. In 2D we have that $\nabla^\perp$ maps
onto the kernal of $\nabla\cdot$, ensuring that
$\nabla\cdot\nabla^\perp(\cdot)=0$. It is possible to write this de
Rham complex such that instead we have that $\nabla$ maps onto the
kernal of $\nabla^\perp$, ensuring that
$\nabla^\perp\cdot\nabla(\cdot)=0$, but the former is more useful in
constructing discretisations for geophysical fluid dynamics since it
enables schemes with local mass conservation \citep{}.

@@

Given that there are many promising avenues for research in the area
of compatible finite element discretisations for geophysical fluid
dynamics (**where to say what these are?), a flexible, extensible
implementation of the methods is strongly desirable. This is
especially required because, while it is feasible for someone to write
a code implementing a specific finite element discretisation of a PDE,
the code quickly becomes complicated and unsustainable when
investigating different orders or families of finite elements, or
different formulations of the equation. For geophysical fluid dynamics
applications there are additional considerations relating to the
geometry and types of waves present in the flow. For example, the
scales of motion in the horizontal and vertical directions differ by
several orders of magnitude. (also: boundaries, vertical staggering
for theta space)

@@

In this article we will describe the current capabilities of Gusto and
the design features that enable our flexible implementation whereby
different spatial and temporal discretisations can be applied to a
range of different equation sets relevant for weather and climate
modelling. In section \ref{sec: governing} we will outline three
different equations commonly used for modelling geophysical fluids in
order to demonstrate the similaries between the equations and justify
the use of a unified codebase. The advantages of this unified approach
become even clearer in section \ref{sec: FEM} where we present the
compatible finite element discretisation of the different types of
terms (e.g. pressure gradient, transport, diffusion) seen in section
\ref{sec: governing}. In section \ref{sec: design} we describe the
code structure of Gusto followed by a detailed example (section
\ref{sec: FML}) of the Form Manipulation Language (FML) that
facilitates symbolic manipulation of the finite element forms within
our timestepping code. Section \ref{sec: results} contains a selection
of numerical results that showcases the range of discretisations
available within Gusto. **mention here new functionality facilitated
by the flexibility of Gusto? We summarise our current status and
future directions in section \ref{sec: summary}.

>>>>

Compatible finite elements %!TB methods?
provide a route to stable, consistent
discretisations of the partial differential equations governing
geophysical fluids %!TB cite acta numerica
%!TB Suggest rewriting and combining the following two sentences:
which also have the correct wave propagation speeds
and avoid spurious computational modes, even on non-uniform
grids.
These properties, along with conservation of mass and
potentially other quantities such as energy and enstrophy, are
considered essential for the accuracy of the dynamical core of weather
and climate models \citep{staniforth2012horizontal}.
Achieving these
properties on a non-orthogonal grid is important because of parallel
scaling**. %!TB citation and expand on this. Reference Met Office and GungHo
While the required properties of the discretisation guide the
choice of finite element spaces, there are many options for both the
spatial and temporal discretisation and it is not clear which will be
best for any particular problem.
%!TB Maybe a bit more direct about what Gusto is and why?
Gusto provides a flexible,
extensible, easy-to-use toolkit of options for rapid prototyping of
different spatial and temporal discretisations based on the compatible
finite element framework applied to geophysical fluid dynamics
equations relevant to numerical weather and climate prediction. In
this paper we introduce Gusto to the community, summarise its current
capabilities and outline our vision for future development.

@@

The finite element method represents fields (e.g. velocity, pressure,
density or temperature) on a mesh made of non-overlapping, conforming
cells, or elements. %!TB: what does conforming mean here? Can we clarify what we mean by element? Do we want to say something about the elements being the same across the mesh?
Within each element, the field is represented as a
sum of coefficients multiplied by polynomial basis functions. The
finite element space is specified by the degree of the polynomial
basis functions and the continuity of these functions between the
elements. The partial differential equations used in weather and
climate modelling have multiple prognostic fields and it is not
necessary, or indeed desirable, to use the same finite element
function space for each field. %!TB: should we distinguish between finite elements and function spaces?
Using different finite element function
spaces for each field gives what is known as a mixed finite element
discretisation.

@@

The compatible finite element spaces relevant for geophysical fluid
dynamics are constructed from a discrete differential complex called
the de Rham complex. This is a sequence of spaces with differential
operators between them that map onto the kernel of the next operator
in the sequence. This is often summarised in a commutation diagram as
in figures \ref{fig:2D de Rham} and \ref{fig:3D de Rham}. In these
diagrams the infinite dimensional spaces are along the top row and the
$\pi_i$ denote projections into the corresponding finite dimensional
subspace used in the compatible finite element discretisation. The
differential operators are denoted by $\diff^k$ and these commute with
the projections, meaning that $\diff^{k+1}\pi_k(\cdot) =
\pi_{k+1}\diff^{k+1}(\cdot)$. This guarantees that, in 3D, $\nabla$
maps onto the kernel of $\nabla\times$, ensuring
$\nabla\times\nabla(\cdot)=0$, and $\nabla\times$ maps onto the kernel
of $\nabla\cdot$, ensuring that $\nabla\cdot\nabla\times(\cdot)=0$;
i.e.\, the spaces are compatible. In 2D we have that $\nabla^\perp$ maps
onto the kernel of $\nabla\cdot$, ensuring that
$\nabla\cdot\nabla^\perp(\cdot)=0$. It is possible to write this de
Rham complex such that instead $\nabla$ maps onto the
kernel of $\nabla^\perp$, ensuring that
$\nabla^\perp\cdot\nabla(\cdot)=0$, but the former is more useful in
constructing discretisations for geophysical fluid dynamics since it
enables schemes with local mass conservation \citep{}. %!TB: citation

@@

Given that there are many promising avenues for research in the area
of compatible finite element discretisations for geophysical fluid
dynamics (**where to say what these are?), a flexible, extensible
implementation of the methods is strongly desirable. %!TB: should we change this sentence to start with the Met Office link? "Because of these promising properties, gungho. However there are still many important avenues for research and unanswered questions
%! - time stepping methods
%! - solver techniques
%! - aspects of physics-dynamics coupling
This is
especially required because, while it is feasible for someone to write
a code implementing a specific finite element discretisation of a PDE,
the code quickly becomes complicated and unsustainable when
investigating different orders or families of finite elements, or
different formulations of the equation. %!TB: this sentence felt a bit clunky. I guess we want to say that "The research of these questions can often be difficult in operational models, which are generally designed to execute specific configurations very efficiently. So the purpose of Gusto is to facilitate research by providing a flexible, extensible framework, in which the implementation of new methods is easier.
For geophysical fluid dynamics
applications there are additional considerations relating to the
geometry and types of waves present in the flow. For example, the
scales of motion in the horizontal and vertical directions differ by
several orders of magnitude. (also: boundaries, vertical staggering
for theta space) %!TB: this point feels a bit out of place here, should we move it above?

@@

In this article we will describe the current capabilities of Gusto and
the design features that enable our flexible implementation whereby
different spatial and temporal discretisations can be applied to a
range of different equation sets relevant for weather and climate
modelling. In section \ref{sec: governing} we will outline three
different equations commonly used for modelling geophysical fluids in
order to demonstrate the similarities between the equations and justify
the use of a unified codebase. The advantages of this unified approach
become even clearer in section \ref{sec: FEM} where we present the
compatible finite element discretisation of the different types of
terms (e.g. pressure gradient, transport, diffusion) seen in section
\ref{sec: governing}. In section \ref{sec: design} we describe the
code structure of Gusto followed by a detailed example (section
\ref{sec: FML}) of the Form Manipulation Language (FML) that
facilitates symbolic manipulation of the finite element forms within
our timestepping code. Section \ref{sec: results} contains a selection
of numerical results that showcases the range of discretisations
available within Gusto. **mention here new functionality facilitated
by the flexibility of Gusto? We summarise our current status and
future directions in section \ref{sec: summary}.
